State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2020-08-26 analysis/paper/paper.Rmd
MA edits to manuscript

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Mike W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  This is the abstract
keywords: |
  escapement; Chinook; steelhead; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

Fish escapement is the number of adults that survive juvenile and adult rearing, escape harvest and return to their natal habitat to potentially spawn (e.g. @Bue1998). For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area. Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for each component group (e.g., stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@mcelhany2000viable], and facilitate assessments of population viability and extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].

Populations of Chinook Salmon, Oncorhynchus tshawytscha, and steelhead trout, O. mykiss, in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015]. As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937). Snake River Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations. Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (**cite BiOp, NPCC FWP, and status assessment**). Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, are predicated on escapements at Lower Granite Dam.

Chinook Salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations. Fish management agencies previously used counts of fish, by species, passing an observation window within the fish ladder as a census of fish escaping past the dam. For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017; @Steinhorst2017]. Window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time. Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making (**citation?**).

Using window counts, however, as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and ignores known biological processes. First, live (i.e., in-person) window counts only occur from April through October each year. Additionally, live window counts only occur for 16-hours each day, and fish counters working at the observation window look directly into the fish ladders to identify and count all passing fish, by species, for only 50 minutes of each hour. Counts are then expanded to provide an estimate for the hour [@USACE2015]. From November through March, the remainder of the year, video tape fish counting is used and only occurs for 10 hours each day; fish counters then read the video tapes and submit daily fish counts [@Hatch1994]. Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored. Additionally, two biological processes are unaccounted for: 1) fish that cross the dam during the 8-hours when the window is closed for counting (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder again [@Boggs2004]. Both fallback with no re-ascension and fallback with re-ascension potentially result in an overestimate of escapement [@Dauble2000]. Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates cancelled each other out resulting in window counts providing an unbiased estimate of escapement (**citation?**)

In this study, we describe a novel method to estimate aggregate and group escapement which incorporates all sources of known uncertainty, we demonstrate the estimation method is essentially unbiased, thus better informing conservation and management decision making, and show night-time passage and fallback/re-ascension rates are unequal. Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of PIT tagged fish in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach [@royle2008hierarchical]. To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are parsed into three origin groups: wild, hatchery, and hatchery no-clip. Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours and account for fish that may ascend the ladder multiple times due to fallback and re-ascension. Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for downloaded at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM). To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates. The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provide more reasonable estimates of uncertainty.<!-- RK comment: the following sentence is better left in the discussion and can be removed from this paragraph. -- Additionally, the STADEM model framework can be applied elsewhere to estimate aggregate and group escapement, provided a passage barrier containing multiple counting and observational mechanisms exist, such as, a counting window, an adult trap sampling a portion of the run, and PIT tag antenna infrastructure. -->

@@

We used STADEM and three sources of data to estimate Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019. Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder. Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; knowing, similar data could likely be obtained from other fish passage facilities.

@@

Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite fish ladder were the first source of data. Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season. Window counts were made for each species using video monitoring and direct in-person visual monitoring during daytime hours [@Hatch1994]. Video monitoring occurred during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) for 10 hours per day (0600 – 1600 hours). Direct visual monitoring occurred during peak run times (April 1 – October 31) for 16 hours per day (0400 – 2000 hours) [@USACE2015]. During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation. Salmonids under 30cm in length were not identified to species. The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted. Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime) [@USACE2015]. Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, [www.cbr.washington.edu/dart/query/adult_daily](www.cbr.washington.edu/dart/query/adult_daily), using their window count query. Counts were provided for each day the fish ladder was open to passage. Although window counts were assumed to be a census of every fish passing Lower Granite Dam, corrections were not applied for nighttime passage or re-ascending fish. Further, there was no estimation of daily or seasonal observation or sampling error.

@@

The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam [@Ogden2016b]. The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., @Camacho2019, @Steinhorst2017). The trap was operational for 24 hours per day, and randomly sampled the daily run by opening four times per hour for a length of time determined by a set daily trapping rate. The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish, but also balancing fish handling concerns. Trap sample rates were typically 10-25%, but fluctuated throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, etc. All captured fish were anesthetized, identified to species, examined for existing marks/tags, and measured for fork length. Additionally, each fish had scale and genetic tissue samples taken; scale samples were used to estimate age [@Wright2015], tissue samples were used to determine sex and for genetic stock identification analysis (e.g., @Hargrove2019). Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, hatchery, and hatchery no-clip origins were assigned using a post-hoc analysis of marks, tags, and genetic information. Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2019].

@@

The last source of data was observations of PIT tagged adult Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder. These observations provide estimates of 1) a trapping rate, 2) the proportion of fish passing during nighttime hours (i.e., outside of window observation hours), and 3) the proportion of fish that ascend the fish ladder multiple times (i.e., re-ascension rate). Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to reaching Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system. PIT tag data was provided through DART and the adult ladder PIT tag query; [http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail](http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail).

A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites. The "mark" group included all tags detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system. Using a mark-recapture model with differing capture probabilities, we estimated the trap rate on a weekly basis. Those estimates, with associated uncertainties, were then incorporated into the model as informed priors, while the model estimated the true trap rate based on all the data, including trap and window counts. <!--The true trapping rate was estimated as the recorded time of the trap being open does not always reflect the true proportion of fish that are captured in the trap, due to trap malfunctions, sort-by-code fish opening the trap more frequently than expected, or process error (Rick Orme, *per. comm.*). Therefore, we use the mark-recapture estimate of trap rate. -->

@@

We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage, and fallback/re-ascension rates using a state-space modeling approach [@royle2008hierarchical] implemented in the STADEM package for the R statistical software [@R-Core-Team2020]. We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error. In the case of the trap for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate. STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder. Finally, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig)). Additional model details can be found in [Appendix A](#append1). The STADEM package is available from the primary author at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.

@@

We tested the STADEM model on a variety of simulated data sets. These simulated data sets contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at Lower Granite Dam). Each simulated fish was given a date of ladder ascension, based on its population and the range of run timing for that population. Each fish was also simulated to cross the dam either while the window was open for counting, or not, and was given the chance to be "caught" in the simulated fish trap given the week when it ascended the dam, and the known trap rate that week. Fallback and re-ascension behavior was also simulated, with each fish having the possibility of falling back and re-ascending the ladder up to three times.

@@

Finally, we applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River to spawn for the spawn years 2010 to 2019. Window counts for both species were accessed from DART via functions within STADEM. For Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall. For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017). Data from the adult trap was made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at Lower Granite Dam was accessed from the PTAGIS regional database ([https://www.ptagis.org/](https://www.ptagis.org/)).

@@

We applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead for spawn years 2010–2019. Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab). Coefficients of variation ranged from 2.7-24.8% for wild fish, 2.2-10.5% for hatchery fish, 3.6-11% for hatchery no-clip fish and 2.1-10.4% for total unique fish past Lower Granite Dam.

<!-- All of the high values were from 2019, and the ranges drop quite a bit if we exclude that year. Max is more like 6%. Worth trying to mention that?  -->

@@

Examining the estimates of nighttime passage and re-ascension rates based on the observed PIT tags crossing over Lower Granite Dam, the two rates did not match up (Figure \@ref(fig:night-reasc-diff-fig)). In particular, there are several weeks when the window counts are quite high, and the rates differ by as much as 10%. Clearly, those two biological processes do not cancel each other out, and thus, employing a model that accounts for both will result in more accurate estimates of escapement.

@@

<!-- I can't find the Steele et al. 2013 reference -->

>>>>

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
- AckermanLibrary.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  This is the abstract
keywords: |
  escapement; Chinook; steelhead; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

Fish escapement often refers to the number of adults that survive juvenile and subadult rearing, escape harvest and return to their natal habitat to potentially spawn (e.g. @Bue1998). For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area. Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for component groups (e.g., by stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017; @Camacho2017) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@mcelhany2000viable], and facilitate assessments of population viability and extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].

Populations of Chinook salmon *Oncorhynchus tshawytscha* and steelhead trout *O. mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015]. As a result, Snake River spring-summer run Chinook salmon (hereafter Chinook salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937). Snake River Chinook salmon and steelhead have substantial cultural, recreational, commercial and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River Chinook salmon and steelhead populations, with the exception of Tucannon River, is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations. Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin [@NMFS2019BiOp; @NFSC2015; @NPCC2014]. Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers fisheries, are predicated on escapements at Lower Granite Dam.

Chinook salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations. Fish management agencies previously used counts of fish, by species, passing an observation window within the fish ladder as a census of fish escaping past the dam. For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017; @Steinhorst2017; @Steele2019]. Window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time. Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making [@WDFW2019].

Using window counts, however, as a census of Chinook salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and disregards known biological processes. First, live (i.e., in-person) window counts only occur from April through October each year. Additionally, live window counts only occur for 16-hours each day, and fish counters working at the observation window look directly into the fish ladders to identify and count all passing fish, by species, for 50 minutes of each hour. Counts are then expanded to provide an estimate for the entire hour [@USACE2015]. From November through March, the remainder of the year, video tape fish counting is used and only occurs for 10 hours each day; fish counters then read the video tapes and submit daily fish counts [@Hatch1994]. Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored. Additionally, two biological processes are unaccounted for: 1) fish that cross the dam during the 8-hours when the window is closed for counting (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder again [@Boggs2004]. Both fallback with no re-ascension and fallback with re-ascension potentially result in an overestimate of escapement [@Dauble2000]. Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates canceled each other out resulting in window counts providing an unbiased estimate of escapement [@Camacho2017].

In this study, we describe a novel method to estimate aggregate and group escapements which incorporate all sources of known uncertainty and demonstrate that the estimation method is essentially unbiased, thus better informing conservation and management decision making. We also show that observed nighttime passage and fallback/re-ascension rates are typically unequal. Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of PIT tagged fish in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach [@royle2008hierarchical]. To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are parsed into three origin groups: wild, hatchery, and hatchery no-clip. Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours and account for fish that may ascend the ladder multiple times due to fallback and re-ascension. Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for downloaded at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM). To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates. The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provides improved estimates of uncertainty.

<!-- RK comment: the following sentence is better left in the discussion and can be removed from this paragraph. -- Additionally, the STADEM model framework can be applied elsewhere to estimate aggregate and group escapement, provided a passage barrier containing multiple counting and observational mechanisms exist, such as, a counting window, an adult trap sampling a portion of the run, and PIT tag antenna infrastructure. -->

@@

We used STADEM and three sources of data to estimate Chinook salmon and steelhead escapement at Lower Granite Dam from 2010-2019. Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder. Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; knowing, similar data could likely be obtained from other fish passage facilities.

@@

Daily counts of adult Chinook salmon and steelhead passing an observation window located on the Lower Granite Dam fish ladder were the first source of data. Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season. Window counts were made for each species using video monitoring and direct in-person visual monitoring during daytime hours [@Hatch1994]. Video monitoring occurred during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) for 10 hours per day (0600 – 1600 hours). Direct visual monitoring occurred during peak run times (April 1 – October 31) for 16 hours per day (0400 – 2000 hours) [@USACE2015]. During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation. Salmonids under 30cm in length were not identified to species. The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted. Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime) [@USACE2015]. Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, [www.cbr.washington.edu/dart/query/adult_daily](www.cbr.washington.edu/dart/query/adult_daily), using their window count query. Counts were provided for each day the fish ladder was open to passage. Although window counts were assumed to be a census of every fish passing Lower Granite Dam, corrections were not applied for nighttime passage or re-ascending fish. Further, there was no estimation of daily or seasonal observation or sampling error.

@@

The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam [@Ogden2016b]. The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., @Camacho2017, @Steinhorst2017). The trap was operational for 24 hours per day, and randomly sampled the daily run by opening four times per hour for a length of time determined by a set daily trapping rate. The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish, but also balancing fish handling concerns. Trap sample rates were typically 10-25%, but fluctuated throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, in-season forecast adjustments, etc.

All captured fish were anesthetized, identified to species, examined for existing marks/tags, and measured for fork length. For adipose-intact (unclipped) adults, which includes wild and hatchery-no clip individuals, some portion or all of fish trapped had scale and genetic tissue samples taken. Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012a] and estimate the origin of wild fish via genetic stock identification (e.g., @Hargrove2019). Starting with adults spawning in 2013, every unclipped Chinook salmon and steelhead trapped at LGR was genotyped to simplify collaborative logistics and better estimate the proportion of phenotypically wild fish of truly hatchery origin. Previous to that, only fish determined to be wild in origin at the trap were sampled for scale and genetics. @Camacho2017 provide further details on trap sample rates and valid sample selection. Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, hatchery, and hatchery no-clip origins were assigned using a post-hoc analysis of marks and tags, including genetic parentage-based tags [@Steele2013; @Steele2019]. Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2017].

@@

The last source of data was observations of PIT tagged adult Chinook salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder. These observations provide estimates of 1) a trapping rate, 2) the proportion of fish passing during nighttime hours, and 3) the proportion of fish that ascend the fish ladder multiple times (i.e., the re-ascension rate). Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to reaching Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system. PIT tag data was provided through DART and the adult ladder PIT tag query; [http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail](http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail).

A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook salmon and steelhead at Lower Granite Dam adult detection sites. The "mark" group included all tags detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system. Using a mark-recapture model with differing capture probabilities, we estimated the trap rate on a weekly basis. Those estimates, with associated uncertainties, were then incorporated into the model as informed priors, while the model estimated the true trap rate based on all the data, including trap and window counts. We want to estimate the "true" trapping rate as the set trapping rate, the recorded time that the trap is open to trap adults, does not always reflect the true proportion of fish that are captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others. Therefore, we use the mark-recapture estimate of the trap rate to estimate a true trapping rate.

<!--The true trapping rate was estimated as the recorded time of the trap being open does not always reflect the true proportion of fish that are captured in the trap, due to trap malfunctions, sort-by-code fish opening the trap more frequently than expected, or process error (Rick Orme, *per. comm.*). Therefore, we use the mark-recapture estimate of trap rate. -->

@@

We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage, and fallback/re-ascension rates using a state-space modeling approach [@royle2008hierarchical] implemented in the STADEM package for the R statistical software [@R-Core-Team2020]. We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error. In the case of the trap, for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate. STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder. Finally, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig)). Additional model details can be found in [Appendix A](#append1). The STADEM package is available from the primary author at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.

@@

We tested the STADEM model on a variety of simulated data sets. These simulated data sets contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at Lower Granite Dam). Each simulated fish was given a date of ladder ascension, based on its population and the range of observed run timing for that population. Each fish was also simulated to cross the dam either while the window was open for counting, or not, and was given the chance to be "caught" in the simulated fish trap given the week when it ascended the dam, and the known trap rate that week. Fallback and re-ascension behavior was also simulated, with each fish having the possibility of falling back and re-ascending the ladder up to three times.

@@

Finally, we applied STADEM to data from Lower Granite Dam for Chinook salmon and steelhead returning to the Snake River during spawn years 2010 to 2019. Window counts for both species were accessed from DART via functions within STADEM. For Chinook salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall. For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017). Data from the adult trap was made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at Lower Granite Dam was accessed from the PTAGIS regional database ([https://www.ptagis.org/](https://www.ptagis.org/)).

@@

We applied STADEM to data from Lower Granite Dam for Chinook salmon and steelhead for spawn years 2010–2019. Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab). Coefficients of variation ranged from 2.7-24.8% for wild fish, 2.2-10.5% for hatchery fish, 3.6-11% for hatchery no-clip fish and 2.1-10.4% for total unique fish past Lower Granite Dam.

<!-- 
All of the high values were from 2019, and the ranges drop quite a bit if we exclude that year. Max is more like 6%. Worth trying to mention that?

MA: Maybe worth including in the Discussion?
-->

@@

Examining the estimates of nighttime passage and re-ascension rates based on the observed PIT tags crossing over Lower Granite Dam (Figure \@ref(fig:night-reasc-diff-fig)), the two rates did not match up in most cases. In particular, there are several weeks when the window counts are quite high, and the rates differ by as much as 10%. Clearly, those two biological processes do not cancel each other out in nearly all cases, and thus, employing a model that accounts for both will result in more accurate estimates of escapement.

@@

Although not currently set up for this, STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers. Currently, the only roadblock to this, at Lower Granite Dam, is the identification of hatchery origin fish from phenotypically wild fish (i.e., hatchery no-clip) using genetic tissue samples [@Steele2013; @Steele2019] collected at the adult trap, which currently is completed post-hoc after the trapping season. The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement. However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement. Final post-hoc estimates, parsed by origin, could then be finalized at season’s end. All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in almost real-time by DART. Provided the Lower Granite Dam adult fish trap database was updated and available in near real-time, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.
